Simple repression draft

### 2020-12-06 doc/resubmission/section_03_beyond_means.tex
remove comment in latex file
final pass before resubmission
minor changes after 2nd round of reviews. Paper is ready

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../../figures/main/fig02.pdf}
\caption{\textbf{Comparison of different models for noise in the constitutive promoter.}
(A) The left column depicts various plausible models for the dynamics of
constitutive promoters. In model (1), transcripts are produced in a Poisson
process~\cite{Sanchez2013, Jones2014}. Model (2) features explicit treatment of
RNAP binding/unbinding kinetics~\cite{Phillips2015a}. Model (3) is a more
detailed generalization of model (2), treating transcription initiation as a
multi-step process proceeding through closed and open
complexes~\cite{Mitarai2015}. Model (4) is somewhat analogous to (2) except with
the precise nature of active and inactive states left
ambiguous~\cite{Peccoud1995, Shahrezaei2008, Razo-Mejia2020}. Finally, model (5)
can be viewed as a certain limit of model (4) in which transcripts are produced
in bursts, and initiation of bursts is a Poisson process. 
% \mmnote{Haven't found model 5 formulated quite like this in the literature,
% usually burstiness is handled like model 4, but I should probably search a bit
% more.} 
The right column shows the Fano factor $\nu$ (variance/mean) for each model.
Note especially the crucial diagnostic: (2) and (3) have $\nu$ strictly below 1,
while only for (4) and (5) can $\nu$ exceed 1. Models with Fano factors $\le 1$
cannot produce the single-cell data observed in part (B) without introducing
additional assumptions and model complexity. (B) Data from \cite{Jones2014}.
Mean mRNA count vs. Fano factor (variance/mean) for different promoters as
determined with single-molecule mRNA Fluorescence \textit{in situ}
Hybridization. The colorbar indicates the predicted binding affinity of RNAP to
the promoter sequence as determined in \cite{Brewster2012}. Numbers serve for
cross comparison with data presented in Figure 3.}
\label{fig2:constit_cartoons}
\end{figure}

@@

Although physically intuitive given our understanding of how a single transcript
is produced, we are forced to discard both of these models on the basis that the
Fano factor for both of them is strictly $< 1$ as shown in
Figure~\ref{fig2:constit_cartoons}(A) (See Appendix~\ref{sec:non_bursty} for
detailed derivations). This is in disagreement with the experimental data
from~\cite{Jones2014}, reproduced in Figure~\ref{fig2:constit_cartoons}(B). All
of these unregulated promoters --which only differ in their promoter sequence--
show strictly super-Poissonian noise (Fano factor $> 1$). In fact, we suspect
\textit{any} model in which transcription proceeds through a multistep cycle
must necessarily have $\nu<1$. The intuitive argument compares the waiting time
distribution to traverse the cycle with the waiting time for a Poisson promoter
(model 1) with the same mean time. The latter is simply an exponential
distribution. The former is a convolution of multiple exponentials, and
intuitively the waiting time distribution for a multistep process should be more
peaked with a smaller fractional width than a single exponential with the same
mean. A less disperse waiting time distribution means transcription initations
are more uniformly distributed in time relative to a Poisson process. Hence the
distribution of mRNA over a population of cells should be less variable compared
to Poisson, giving $\nu<1$. (In Appendix~\ref{sec:non_bursty} we present a more
precise version of the intuitive arguments in this paragraph).

@@

Consider the time delay between the first and final RNAPs in a burst initiating
transcription (\textit{not} the time to complete transcripts, which potentially
could be much longer.) If this timescale is short compared to the typical search
timescale of transcription factors, then all is well. The estimates from
deHaseth et.\ al.~\cite{DeHaseth1998} put RNAP's diffusion-limited on rate
around $\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$ and polymerase
loading as high as $1~\text{s}^{-1}$. Then for reasonable burst sizes of $<10$,
it is reasonable to guess that bursts might finish initiating on a timescale of
tens of seconds or less (with another 30-60 sec to finish elongation, but that
does not matter here). A transcription factor with typical copy number of order
10 (or less) would have a diffusion-limited association rate of order
$(10~\text{sec})^{-1}$ \cite{Hammar2014}. Higher copy number TFs tend to have
many binding sites over the genome, which should serve to pull them out of
circulation and keep their effective association rates from rising too large.
Therefore, there is \textit{perhaps} a timescale separation possible between
transcription factor association rates and burst durations, but this assumption
could very well break down, so we will have to keep it in mind when we infer
repressor rates from the Jones et.\ al.\ single-cell mRNA counts data
later~\cite{Jones2014}.

>>>>

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../../figures/main/fig02.pdf}
\caption{\textbf{Comparison of different models for noise in the constitutive promoter.}
(A) The left column depicts various plausible models for the dynamics of
constitutive promoters. In model (1), transcripts are produced in a Poisson
process~\cite{Sanchez2013, Jones2014}. Model (2) features explicit treatment of
RNAP binding/unbinding kinetics~\cite{Phillips2015a}. Model (3) is a more
detailed generalization of model (2), treating transcription initiation as a
multi-step process proceeding through closed and open
complexes~\cite{Mitarai2015}. Model (4) is somewhat analogous to (2) except with
the precise nature of active and inactive states left
ambiguous~\cite{Peccoud1995, Shahrezaei2008, Razo-Mejia2020}. Finally, model (5)
can be viewed as a certain limit of model (4) in which transcripts are produced
in bursts, and initiation of bursts is a Poisson process. The right column shows
the Fano factor $\nu$ (variance/mean) for each model. Note especially the
crucial diagnostic: (2) and (3) have $\nu$ strictly below 1, while only for (4)
and (5) can $\nu$ exceed 1. Models with Fano factors $\le 1$ cannot produce the
single-cell data observed in part (B) without introducing additional assumptions
and model complexity. (B) Data from \cite{Jones2014}. Mean mRNA count vs. Fano
factor (variance/mean) for different promoters as determined with
single-molecule mRNA Fluorescence \textit{in situ} Hybridization. The colorbar
indicates the predicted binding affinity of RNAP to the promoter sequence as
determined in \cite{Brewster2012}. Numbers serve for cross comparison with data
presented in Figure 3.}
\label{fig2:constit_cartoons}
\end{figure}

@@

Although physically intuitive given our understanding of how a single transcript
is produced, we are forced to discard both of these models on the basis that the
Fano factor for both of them is strictly $< 1$ as shown in
Figure~\ref{fig2:constit_cartoons}(A) (See Appendix~\ref{sec:non_bursty} for
detailed derivations). This is in disagreement with the experimental data
from~\cite{Jones2014}, reproduced in Figure~\ref{fig2:constit_cartoons}(B). All
of these unregulated promoters --which only differ in their promoter sequence--
show strictly super-Poissonian noise (Fano factor $> 1$). In fact, we suspect
\textit{any} model in which transcription proceeds through a multistep cycle
must necessarily have $\nu<1$. The intuitive argument compares the waiting time
distribution to traverse the cycle with the waiting time for a Poisson promoter
(model 1) with the same mean time. The latter is simply an exponential
distribution. The former is a convolution of multiple exponentials, and
intuitively the waiting time distribution for a multistep process should be more
peaked with a smaller fractional width than a single exponential with the same
mean. A less disperse waiting time distribution means transcription initiations
are more uniformly distributed in time relative to a Poisson process. Hence the
distribution of mRNA over a population of cells should be less variable compared
to Poisson, giving $\nu<1$. (In Appendix~\ref{sec:non_bursty} we present a more
precise version of the intuitive arguments in this paragraph).

@@

Consider the time delay between the first and final RNAPs in a burst initiating
transcription (\textit{not} the time to complete transcripts, which potentially
could be much longer.) If this timescale is short compared to the typical search
timescale of transcription factors, then all is well. The estimates from
deHaseth et.\ al.~\cite{DeHaseth1998} put RNAP's diffusion-limited on rate
around $\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$ and polymerase
loading as high as $1~\text{s}^{-1}$. Then for reasonable burst sizes of $<10$,
it is reasonable to guess that bursts might finish initiating on a timescale of
tens of seconds or less (with another 30-60 sec to finish elongation, but that
does not matter here). A transcription factor with typical copy number of order
10 (or less) would have a diffusion-limited association rate of order
$(10~\text{sec})^{-1}$ \cite{Hammar2014}. Higher copy number TFs tend to have
many binding sites over the genome, which should serve to pull them out of
circulation and keep their effective association rates from rising too
large~\cite{Schmidt2016}. Therefore, there is \textit{perhaps} a timescale
separation possible between transcription factor association rates and burst
durations, but this assumption could very well break down, so we will have to
keep it in mind when we infer repressor rates from the Jones et.\ al.\
single-cell mRNA counts data later~\cite{Jones2014}.

### 2020-12-06 doc/resubmission/section_04_bayesian_inference.tex
minor changes after 2nd round of reviews. Paper is ready

Let us now consider the problem of parameter inference for model five
from~\fig{fig1:means_cartoons}(C). As derived in
Appendix~\ref{sec:gen_fcn_appdx}, the steady-state mRNA distribution in this
model is a negative binomial distribution, given by
\begin{equation}
p(m) = \frac{\Gamma(m+k_i)}{\Gamma(m+1)\Gamma(k_i)}
        \left(\frac{1}{1+b}\right)^{k_i}
        \left(\frac{b}{1+b}\right)^m,
\label{eq:neg_bionom}
\end{equation}
where $b$ is the mean burst size and $k_i$ is the burst rate in units of the
mRNA degradation rate $\gamma$. As sketched earlier, to think of the negative
binomial distribution in terms of an intuitive ``story,'' in the precise
meaning of~\cite{Blitzstein2015}, we imagine the arrival of
bursts as a Poisson process with rate $k_i$, where each burst has a
geometrically-distributed size with mean size $b$.

@@

On the whole then, we regard this as striking confirmation of the validity of
the thermodynamic models. Their lynchpin parameter is a phenomenological free
energy of repressor binding that has previously only been inferred indirectly.
Our result shows that the microscopic interpretation of this free energy, as the
log of a ratio of transition rates, does indeed hold true to within the inherent
uncertainties that remain in the entire theory-experiment dialogue.

>>>>

Let us now consider the problem of parameter inference for model five
from~\fig{fig1:means_cartoons}(C). As derived in
Appendix~\ref{sec:gen_fcn_appdx}, the steady-state mRNA distribution in this
model is a negative binomial distribution~\cite{Kumar2015}, given by
\begin{equation}
p(m) = \frac{\Gamma(m+k_i)}{\Gamma(m+1)\Gamma(k_i)}
        \left(\frac{1}{1+b}\right)^{k_i}
        \left(\frac{b}{1+b}\right)^m,
\label{eq:neg_bionom}
\end{equation}
where $b$ is the mean burst size and $k_i$ is the burst rate in units of the
mRNA degradation rate $\gamma$. As sketched earlier, to think of the negative
binomial distribution in terms of an intuitive ``story,'' in the precise
meaning of~\cite{Blitzstein2015}, we imagine the arrival of
bursts as a Poisson process with rate $k_i$, where each burst has a
geometrically-distributed size with mean size $b$.

@@

On the whole then, we find these to be results very suggestive. Although not
conclusively, our results point at the possible interpretation of the
phenomenological free energy of transcription factor binding as the log of a
ratio of transition rates. This did not need to be the case; the free energy was
inferred from bulk measurements of mean gene expression relative to an
unregulated strain --the fold-change-- while the rates were extracted from
fitting single-molecule counts to the full mRNA steady state distribution.
Further single molecule experiments along the lines of~\cite{Hammar2014} could
shed further light into this idea.
% On the whole then, we regard this as striking confirmation of the validity of
% the thermodynamic models. Their lynchpin parameter is a phenomenological free
% energy of repressor binding that has previously only been inferred indirectly.
% Our result shows that the microscopic interpretation of this free energy, as the
% log of a ratio of transition rates, does indeed hold true to within the inherent
% uncertainties that remain in the entire theory-experiment dialogue.

### 2020-12-06 doc/resubmission/section_05_discussion.tex
final pass before resubmission
minor changes after 2nd round of reviews. Paper is ready

The present work has tackled that question in the context of the first
regulatory architecture hypothesized in the molecular biology era, namely, the
repressor-operator model of Jacob and Monod~\cite{Jacob1961}. Regulation in that
architecture is the result of a competition between a repressor which inhibits
transcription and RNAP polymerase which undertakes it. Through the labors of
generations of geneticists, molecular biologists and biochemists, an
overwhelming amount of information and insight has been garnered into this
simple regulatory motif, licensing it as what one might call the ``hydrogen
atom'' of regulatory biology. It is from that perspective that the present paper
explores the extent to which some of the different models that have been
articulated to describe that motif allow us to understand both the average level
of gene expression found in a population of cells, the intrinsic cell-to-cell
variability, and the full gene expression distribution found in such a population
as would be reported in a single molecule mRNA Fluorescence \textit{in situ}
Hybridization experiment, for example.

@@

It is important to be clear on the nature of the assumptions  present in all
models, including model 5 from Figure~\ref{fig2:constit_cartoons}(A). First, as
with many of the models used routinely in the analysis of transcription, the
models considered here imagine the promoter as existing in a series of discrete
states. This assumption is clearly an oversimplification due to effects such as
DNA supercoiling~\cite{Chong2014,Sevier2016} or DNA
looping~\cite{Boedicker2013a} which are both parameterized by  continuous
variables describing DNA shape; although there are formal and very interesting
ways of integrating out such degrees of freedom. Another facet of the models
used here is that they can feature irreversible processes characterized by only
a single rate process with no corresponding return pathway at odds with
thermodynamics. These irreversible reactions lead to physical paradoxes such as
infinite entropy production rates. Such models are mathematically convenient and
serve only as an approximation to the more realistic situation in which the
forward rates are much larger than the backward rates. One consequence is that
the inferred rates for the transcription burst rate and burst size  can only be
thought of as effective rate constants. Another idealization of the models
considered here is that when computing the fold-change in gene expression
(Section 2) we assume that the translation efficiency of an mRNA transcript is
the same in cells with and without regulation. There might be some known
biological effects that could limit the applicability of this assumption such as
cooperation between the transcription and translation
machinery~\cite{Proshkin2010}, or coupling of mRNA transcription and
degradation~\cite{Chen2015} among other potential effects. Another implicit
assumption in all of the models discussed here is that the effects due to
changes in gene copy number during the cell cycle have no effects on the
expression profile. For moderate growth rates such as those used
in~\cite{Jones2014} from which we obtained the data, at all points during the
cell cycle there are at most two copies of the gene of interest. In a recent
publication, we explored the effects that this change in gene dosage can have in
gene expression, finding that the effects are much more drastic at the protein
level~\cite{Razo-Mejia2020}.

>>>>

The present work has tackled that question in the context of the first
regulatory architecture hypothesized in the molecular biology era, namely, the
repressor-operator model of Jacob and Monod~\cite{Jacob1961}. Regulation in that
architecture is the result of a competition between a repressor which inhibits
transcription and RNAP polymerase which undertakes it. Through the labors of
generations of geneticists, molecular biologists and biochemists, an
overwhelming amount of information and insight has been garnered into this
simple regulatory motif, licensing it as what one might call the ``hydrogen
atom'' of regulatory biology. It is from that perspective that the present paper
explores the extent to which some of the different models that have been
articulated to describe that motif allow us to understand both the average level
of gene expression found in a population of cells, the intrinsic cell-to-cell
variability, and the full gene expression distribution found in such a population
as would be reported in a single molecule mRNA Fluorescence \textit{in situ}
Hybridization experiment, for example.
We do so by contrasting two theoretical frameworks to think about the problem.
On the one hand, thermodynamic models are convenient ways to describe the
problem of gene regulation because of the relatively small number of parameters
compared with kinetic models, and the agnostic nature with respect to the full
reaction topology that leads to a transcription event. In addition, either
implicitly or explicitly, many approaches to gene regulation focus strictly on
occupancy and ideas such as $K_d$s, and thus is of great interest to test these
models. This comes with the limitation that thermodynamic models are only able
to make statements about mean gene expression of a quasi-steady state process.
On the other hand, kinetic models break those limitations, being able to make
explicit predictions about higher moments of the gene expression distribution,
and the transient states towards steady state, at the cost of propagating the
number of parameters, and demanding the full set of reactions

@@

It is important to be clear on the nature of the assumptions  present in all
models, including model 5 from Figure~\ref{fig2:constit_cartoons}(A). First, as
with many of the models used routinely in the analysis of transcription, the
models considered here imagine the promoter as existing in a series of discrete
states. This assumption is clearly an oversimplification due to effects such as
DNA supercoiling~\cite{Chong2014,Sevier2016} or DNA
looping~\cite{Boedicker2013a} which are both parameterized by  continuous
variables describing DNA shape; although there are formal and very interesting
ways of integrating out such degrees of freedom. Another facet of the models
used here is that they can feature irreversible processes characterized by only
a single rate process with no corresponding return pathway at odds with
thermodynamics. Such models are mathematically convenient and serve only as an
approximation to the more realistic situation in which the forward rates are
much larger than the backward rates. One consequence is that the inferred rates
for the transcription burst rate and burst size  can only be thought of as
effective rate constants. Another idealization of the models considered here is
that when computing the fold-change in gene expression (Section 2) we assume
that the translation efficiency of an mRNA transcript is the same in cells with
and without regulation. There might be some known biological effects that could
limit the applicability of this assumption such as cooperation between the
transcription and translation machinery~\cite{Proshkin2010}, or coupling of mRNA
transcription and degradation~\cite{Chen2015} among other potential effects.
Another implicit assumption in all of the models discussed here is that the
effects due to changes in gene copy number during the cell cycle have no effects
on the expression profile. For moderate growth rates such as those used
in~\cite{Jones2014} from which we obtained the data, at all points during the
cell cycle there are at most two copies of the gene of interest. In a recent
publication, we explored the effects that this change in gene dosage can have in
gene expression, finding that the effects are much more drastic at the protein
level~\cite{Razo-Mejia2020}.
